Enforced Invariant Confluence

### 2018-04-22 doc/paper/sections/evaluation.tex
Finished up a TODO in the eval section.
Shrunk eval.

Lucy also includes an implementation of the invariant-confluence and
segmented-invariant confluence system models in roughly 3,500 lines of C++.
Users specify objects, transactions, invariants, and segmentations in C++. Lucy
then replicates the objects using segmented invariant-confluence (or
invariant-confluence if the segmentation contains a single segment without any
disallowed transactions). Clients send every transaction request to a randomly
selected server. When a server receives a transaction request, it executes
\algoref{TxnExecution} to attempt to execute the transaction locally. If the
transaction requires global coordination, then the server forwards the
transaction request to a predetermined leader. When the leader receives a
transaction request, it sends a coordination request to all other servers. When
a server receives a coordination request from the leader, it stops processing
transactions and sends the leader its state in a coordination reply. When the
leader receives coordination replies from all other servers, it executes the
transaction, and then sends its state to the other servers. When a server
receives a new state, it adopts the state, computes its new active segment, and
resumes normal processing. After every 100 transactions processed, a server
sends a merge request to a randomly selected server. Merge requests are tagged
with a monotonically increasing epoch number that is incremented by the master
after every round of global coordination. This allows servers to discard merge
requests from previous epochs.

Lucy can also replicate an object with eventual consistency and with
linearizability. With eventual consistency, clients send every transaction
request to a randomly selected server. The server executes the transaction
locally and returns immediately to the client, sending merge requests after
every 100 transactions. With linearizability, clients send every transaction
request to a predetermined leader. The leader relays the transaction request to
all other servers, and when it receives replies from them, it executes the
transaction and replies to the client. This communication pattern mimics the
``normal operation'' of state machine replication protocols like
Paxos~\cite{lamport1998part} and Viewstamped
Replicated~\cite{liskov2012viewstamped}. In
\secref{SegmentedInvariantConfluenceEval}, we compare the performance of
replicating with segmented invariant-confluence against the performance of
replicating with eventual consistency and linearizability.

Because fault-tolerance is largely an orthogonal concern to
invariant-confluence, Lucy is implemented without fault-tolerance. It would be
straightforward to add fault-tolerance to Lucy, but it would not affect our
discussions or evaluation, so we leave it for future work. Moreover, users
currently have to specify their workloads in Python (for the decision
procedures) and C++ (for the runtime). In the future, we plan on removing this
redundancy.

\subsection{Decision Procedures}
We now evaluate the practicality and efficiency of our decision procedure
prototypes. We begin by demonstrating the decision procedure on a handful of
simple, yet practical examples. We then, discuss TODO

% \todo{Bank account with and without decrements.}
% \todo{Auction example.}
% \todo{Foreign keys with and without deletions.}
% \subsection{TPC and Feral}
% \todo{See which of the Feral and TPC transactions we can determine automatically}
\example[$\ints^2$]\examplelabel{TwoIntsEval}
We begin with a minimal working example. Consider again our recurring example
of $\ints^2$ from \exampleref{Z2}. The Python code used to describe the object,
transactions, and invariant is given in \figref{Z2Code}. When we call
\python{checker.check()}, the interactive decision procedure produces
a counterexample in less than a tenth of a second.  After we label the
counterexample and refine the invariant with $y \leq 0$, the interactive
decision procedure determines that the object is invariant-confluent, again, in
less than a tenth of a second. Note that the object is invariant-confluent but
\emph{not} invariant-closed, so prior work that relies on invariant-closure
alone to determine invariant-confluence would not be able to identify this
example as invariant-confluent.

@@

We can use 2P-sets to model a simple relational database with foreign key
constraints. Let object $O = (X, Y) = ((A_X, R_X), (A_Y, R_Y))$ consist of a
pair of two 2P-Sets $X$ and $Y$, which we view as relations. Our invariant $X
\subseteq Y$ (i.e. $(A_X - R_X) \subseteq (A_Y - R_Y)$) models a foreign key
constraint from $X$ to $Y$. We ran our decision procedure on the object with
initial state $((\emptyset, \emptyset), (\emptyset, \emptyset))$ and
transactions that allow arbitrary insertions and deletions into $X$ and $Y$.
After less than a tenth of a second, the decision procedure produced a
reachable counterexample witnessing the fact that the object is not
invariant-confluent. A concurrent insertion into $X$ and deletion from $Y$ can
lead to a state which violates the invariant. This object is not
invariant-confluent and therefore not invariant-closed. Thus, previous tools
depending on invariant-closure as a sufficient condition would be unable to
conclude definitively that the object is \emph{not} invariant-confluent.

@@

\example[Auction]\examplelabel{AuctionEval}
We now consider a simple auction system introduced in~\cite{gotsman2016cause}.
Our object consists of a set $B$ of integer-valued bids and a optional winning
bid $w$. Initially, $B = \emptyset$ and $w = \bot$ (indicating that there is no
winning bid yet) and we merge states by taking the union of $B$ and the maximum
of $w$ (where $\bot < n$ for all integers $n$). One transaction $t_b$ places a bid
$b$ by inserting it into $B$. Another transaction $t_\text{close}$ closes the
auction and sets $w$ equal to the largest bid in $B$. Our invariant is that if
the auction is closed (i.e.\ $w \neq \bot$), then $w = \max(B)$. We ran our
decision procedure on this example and in a third of a second, it produced a
reachable counterexample witnessing the fact that the object is not
invariant-confluent.  If we concurrently close the auction and place an large
bid, then we can end up in a state in which the auction is closed, but there is
a higher bid in $B$.

We then segmented our object as follows. The first segment $(\setst{(B, w)}{w =
\bot}, \setst{t_b}{b \in \ints})$ allows any bidding transaction so long as the
bid is open. The second segment $(\setst{B, w}{I \cap w \neq \bot}, \emptyset)$
includes all auctions which have already been closed and forbids all
transactions. This segmentation captures the intuition that bids should be
permitted only when the auction is open. We ran our segmented
invariant-confluence decision procedure on this example, and it was able to
deduce without any human interaction that the example was segmented
invariant-confluent in less than a tenth of a second.

\example[Escrow Transactions]\examplelabel{EscrowTransactionsEval}
Escrow transactions are a concurrency control technique that allows a database
to execute transactions that increment and decrement numeric values with more
concurrency than is otherwise possible with general-purpose techniques like
two-phase locking~\cite{o1986escrow}. The main idea is that a portion of the
numeric value is put in escrow, after which a transaction can freely decrement
the value so long as it is not decremented by more than the amount that has
been escrowed. We now show how segmented invariant-confluence can be used to
implement escrow transactions.

@@

\example[TPC-C]\examplelabel{TpccEval}
TPC-C is a ubiquitous OLTP database benchmark with a workload that models a
simple warehousing application~\cite{difallah2013oltp}. The TPC-C specification
outlines twelve ``consistency requirements'' (read invariants) that govern the
warehousing application. In~\cite{bailis2014coordination}, Bailis et al.\
categorize the invariants into one of three types:
\begin{enumerate}
  \item
    Three of the twelve invariants are foreign key constraints.  As discussed
    in \exampleref{ForeignKeysEval}, our decision procedures can automatically
    verify conditions under which foreign key constraints are
    invariant-confluent.

  \item
    \newcommand{\ttt}[1]{{\smaller \texttt{#1}}}
    Seven of the twelve invariants involve maintaining arithmetic relationships
    between relations. Our decision procedures can correctly identify these as
    invariant-confluent. Consider, for example, invariant 1 which dictates that
    a warehouse's year to date balance \ttt{W\_YTD} is equal to the sum of the
    district year to date balances \ttt{D\_YTD} of the twenty districts that
    are associated with the warehouse. The Payment transaction randomly selects
    a district and increments \ttt{W\_YTD} and \ttt{D\_YTD} by a randomly
    generated amount. We model this workload with a PN-Counter for \ttt{W\_YTD}
    and twenty PN-Counters for the twenty instances of \ttt{D\_YTD}. We applied
    Lucy to this workload, and it determined that the workload was
    invariant-confluent in less than a second.

  \item
    Two of the twelve invariants involve generating sequential and unique
    identifiers. This workload is \emph{not} invariant-confluent, but when the
    sequentiality requirement is dropped, it is invariant-confluent.
    Unfortunately, modelling unique key generation in our theoretical framework
    (and in our prototype implementation) is not possible because we have thus
    far tacitly assumed that transaction execution is deterministic. We leave a
    generalization of our theory to non-deterministic transactions for future
    work.
\end{enumerate}

\subsection{Segmented Invariant Confluence}%
\seclabel{SegmentedInvariantConfluenceEval}
Now, we evaluate the performance of replicating an object with segmented
invariant-confluence as compared to replicating it with eventual consistency or
linearizability. We begin with two benchmarks that demonstrate the same
concept: the performance of segmented invariant-confluent replication varies
with the amount of global coordination induced by either (a) performing a
transaction that is disallowed within a segment or (b) transitioning between
segments.

@@

Both eventually consistent replication and linearizable replication are
unaffected by the workload, achieving roughly 700,000 and 7,000 transactions
per second respectively. Expectedly, eventually consistent replication
significantly outperforms linearizable replication because (a) transactions can
be sent to any server (not just the leader) and (b) servers do not coordinate
with each other at all.
%
Segmented invariant-confluent replication performs well for low-withdrawal
workloads and performs increasingly poorly as we increase the fraction of
withdrawal transactions, eventually becoming slower that linearizable
replication. For example, with 5\% withdrawal transactions, segmented
invariant-confluent replication performs an order of magnitude better than
linearizable replication; with 50\% withdrawals, it performs as well; and with
100\% withdrawals, it performs two times worse.

@@

  Eventually consistent replication scales perfectly with the number of nodes,
  confirming the results in~\cite{bailis2014coordination}. With eventually
  consistent replication, servers do not coordinate at all, so they are
  completely unaffected by the number of servers. Linearizable replication, on
  the other hand, scales up to about 3-5 servers before performance begins to
  decrease. These numbers are consistent with typical deployments of
  state-machine replication protocols like Paxos~\cite{chandra2007paxos}.
  Because all messages are sent to the leader, the leader becomes the
  bottleneck as the number of servers and clients increases. Moreover, the
  leader must wait for responses from more servers, increasing the latency of
  the slowest response which in turn decreases throughput.
  Segmented invariant-confluent replication scales up to about 6-8 servers
  before succumbing to the same scalability bottlenecks as linearizable
  replication.

>>>>

Lucy also includes an implementation of the invariant-confluence and
segmented-invariant confluence system models in roughly 3,500 lines of C++.
Users specify objects, transactions, invariants, and segmentations in C++. Lucy
then replicates the objects using segmented invariant-confluence (or
invariant-confluence if the segmentation contains a single segment without any
disallowed transactions). Clients send every transaction request to a randomly
selected server. When a server receives a transaction request, it executes
\algoref{TxnExecution} to attempt to execute the transaction locally. If the
transaction requires global coordination, then the server forwards the
transaction request to a predetermined leader. When the leader receives a
transaction request, it sends a coordination request to all other servers. When
a server receives a coordination request from the leader, it stops processing
transactions and sends the leader its state in a coordination reply. When the
leader receives coordination replies from all other servers, it executes the
transaction, and then sends its state to the other servers. When a server
receives a new state, it adopts the state, computes its new active segment, and
resumes normal processing. After every 100 transactions processed, a server
sends a merge request to a randomly selected server.
% Merge requests are tagged with a monotonically increasing epoch number that is
% incremented by the master after every round of global coordination. This allows
% servers to discard merge requests from previous epochs.

Lucy can also replicate an object with eventual consistency and with
linearizability. With eventual consistency, clients send every transaction
request to a randomly selected server. The server executes the transaction
locally and returns immediately to the client, sending merge requests after
every 100 transactions. With linearizability, clients send every transaction
request to a predetermined leader. The leader relays the transaction request to
all other servers, and when the leader receives replies from them, it executes
the transaction and replies to the client. This communication pattern mimics
the ``normal operation'' of state machine replication protocols
\cite{lamport1998part, liskov2012viewstamped}.
% In \secref{SegmentedInvariantConfluenceEval}, we compare the performance of
% replicating with segmented invariant-confluence against the performance of
% replicating with eventual consistency and linearizability.

Because fault-tolerance is largely an orthogonal concern to
invariant-confluence, Lucy is implemented without fault-tolerance. It would be
straightforward to add fault-tolerance to Lucy, but it would not affect our
discussions or evaluation, so we leave it for future work.
% Moreover, users currently have to specify their workloads in Python (for the
% decision procedures) and C++ (for the runtime). In the future, we plan on
% removing this redundancy.

\subsection{Decision Procedures}
We now evaluate the practicality and efficiency of our decision procedure
prototypes. We begin by demonstrating the decision procedure on a handful of
simple, yet practical examples. We then discuss how our tool can be used to
analyze the TPC-C benchmark.

\example[$\ints^2$]\examplelabel{TwoIntsEval}
We begin with a minimal working example. Consider again our recurring example
of $\ints^2$ from \exampleref{Z2}. The Python code used to describe the object,
transactions, and invariant is given in \figref{Z2Code}. When we call
\python{checker.check()}, the interactive decision procedure produces a
counterexample in less than a tenth of a second.  After we label the
counterexample and refine the invariant with $y \leq 0$, the interactive
decision procedure determines that the object is invariant-confluent, again, in
less than a tenth of a second. Note that the object is invariant-confluent but
\emph{not} invariant-closed, so prior work that relies on invariant-closure
alone to determine invariant-confluence would not be able to identify this
example as invariant-confluent.

@@

We can use 2P-sets to model a simple relational database with foreign key
constraints. Let object $O = (X, Y) = ((A_X, R_X), (A_Y, R_Y))$ consist of a
pair of two 2P-Sets $X$ and $Y$, which we view as relations. Our invariant $X
\subseteq Y$ (i.e. $(A_X - R_X) \subseteq (A_Y - R_Y)$) models a foreign key
constraint from $X$ to $Y$. We ran our decision procedure on the object with
initial state $((\emptyset, \emptyset), (\emptyset, \emptyset))$ and
with transactions that allow arbitrary insertions and deletions into $X$ and
$Y$. After less than a tenth of a second, the decision procedure produced a
reachable counterexample witnessing the fact that the object is not
invariant-confluent. A concurrent insertion into $X$ and deletion from $Y$ can
lead to a state which violates the invariant. This object is not
invariant-confluent and therefore not invariant-closed. Thus, previous tools
depending on invariant-closure as a sufficient condition would be unable to
conclude definitively that the object is \emph{not} invariant-confluent.

@@

\example[Auction]\examplelabel{AuctionEval}
We now consider a simple auction system introduced in~\cite{gotsman2016cause}.
Our object consists of a set $B$ of integer-valued bids and a optional winning
bid $w$. Initially, $B = \emptyset$ and $w = \bot$ (indicating that there is no
winning bid yet) and we merge states by taking the union of $B$ and the maximum
of $w$ (where $\bot < n$ for all integers $n$). One transaction $t_b$ places a bid
$b$ by inserting it into $B$. Another transaction $t_\text{close}$ closes the
auction and sets $w$ equal to the largest bid in $B$. Our invariant is that if
the auction is closed (i.e.\ $w \neq \bot$), then $w = \max(B)$. We ran our
decision procedure on this example and in a third of a second, it produced a
reachable counterexample witnessing the fact that the object is not
invariant-confluent.  If we concurrently close the auction and place a large
bid, then we can end up in a state in which the auction is closed, but there is
a bid in $B$ larger than $w$.

We then segmented our object as follows. The first segment $(\setst{(B, w)}{w =
\bot}, \setst{t_b}{b \in \ints})$ allows bidding so long as the bid is open.
The second segment $(\setst{B, w}{w \neq \bot} \cap I, \emptyset)$ includes all
auctions which have already been closed and forbids all transactions. This
segmentation captures the intuition that bids should be permitted only when the
auction is open. We ran our segmented invariant-confluence decision procedure
on this example, and it was able to deduce without any human interaction that
the example was segmented invariant-confluent in less than a tenth of a second.

\example[Escrow Transactions]\examplelabel{EscrowTransactionsEval}
Escrow transactions are a concurrency control technique that allows a database
to execute transactions that increment and decrement numeric values with more
concurrency than is otherwise possible with general-purpose techniques like
two-phase locking~\cite{o1986escrow}. The main idea is that a portion of the
numeric value is put in escrow, after which a transaction can freely decrement
the value so long as it is not decremented by more than the amount that has
been escrowed. We show how segmented invariant-confluence can be used to
implement escrow transactions.

@@

\example[TPC-C]\examplelabel{TpccEval}
TPC-C is a ubiquitous OLTP database benchmark with a workload that models a
simple warehousing application~\cite{difallah2013oltp}. The TPC-C specification
outlines twelve ``consistency requirements'' (read invariants) that govern the
warehousing application. In~\cite{bailis2014coordination}, Bailis et al.\
categorize the invariants into one of three types:
\begin{enumerate}
  \item
    Three of the twelve invariants are \textbf{foreign key constraints}.  As
    discussed in \exampleref{ForeignKeysEval}, our decision procedures can
    automatically verify conditions under which foreign key constraints are
    invariant-confluent.

  \item
    \newcommand{\ttt}[1]{{\smaller \texttt{#1}}}
    Seven of the twelve invariants involve \textbf{maintaining arithmetic
    relationships between relations}. Our decision procedures can correctly
    identify these as invariant-confluent. Consider, for example, invariant 1
    which dictates that a warehouse's year to date balance \ttt{W\_YTD} is
    equal to the sum of the district year to date balances \ttt{D\_YTD} of the
    twenty districts that are associated with the warehouse. The Payment
    transaction randomly selects a district and increments \ttt{W\_YTD} and
    \ttt{D\_YTD} by a randomly generated amount. We model this workload with a
    PN-Counter for \ttt{W\_YTD} and twenty PN-Counters for the twenty instances
    of \ttt{D\_YTD}. We applied Lucy to this workload, and it determined that
    the workload was invariant-confluent in less than a second.

  \item
    Two of the twelve invariants involve generating \textbf{sequential and
    unique identifiers}. This workload is \emph{not} invariant-confluent, but
    when the sequentiality requirement is dropped, it is invariant-confluent.
    Unfortunately, modelling unique key generation in our theoretical framework
    (and in our prototype implementation) is not possible because we have thus
    far tacitly assumed that transaction execution is deterministic. We leave a
    generalization of our theory to non-deterministic transactions for future
    work.
\end{enumerate}

\subsection{Segmented Invariant Confluence}%
\seclabel{SegmentedInvariantConfluenceEval}
Now, we evaluate the performance of replicating an object with segmented
invariant-confluence as compared to replicating it with eventual consistency or
linearizability.
% We begin with two benchmarks that demonstrate the same concept: the performance
% of segmented invariant-confluent replication varies with the amount of global
% coordination induced by either (a) performing a transaction that is disallowed
% within a segment or (b) transitioning between segments.

@@

Both eventually consistent replication and linearizable replication are
unaffected by the workload, achieving roughly 700,000 and 7,000 transactions
per second respectively.
%
% Expectedly, eventually consistent replication significantly outperforms
% linearizable replication because (a) transactions can be sent to any server
% (not just the leader) and (b) servers do not coordinate with each other at all.
%
Segmented invariant-confluent replication performs well for low-withdrawal
workloads and performs increasingly poorly as we increase the fraction of
withdrawal transactions, eventually performing worse than linearizable
replication. For example, with 5\% withdrawal transactions, segmented
invariant-confluent replication performs an order of magnitude better than
linearizable replication; with 50\% withdrawals, it performs as well; and with
100\% withdrawals, it performs two times worse.

@@

  Eventually consistent replication scales perfectly with the number of nodes,
  confirming the results in~\cite{bailis2014coordination}.
  %
  % With eventually consistent replication, servers do not coordinate at all, so
  % they are completely unaffected by the number of servers.
  %
  Linearizable replication, on the other hand, scales up to about 3-5 servers
  before performance begins to decrease. These numbers are consistent with
  typical deployments of state-machine replication protocols like
  Paxos~\cite{chandra2007paxos}.
  %
  % Because all messages are sent to the leader, the leader becomes the
  % bottleneck as the number of servers and clients increases. Moreover, the
  % leader must wait for responses from more servers, increasing the latency of
  % the slowest response which in turn decreases throughput.
  %
  Segmented invariant-confluent replication scales up to about 6-8 servers
  before succumbing to the same scalability bottlenecks as linearizable
  replication.
